Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-11-18 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Rearranges figures to avoid filling a page
Fixing typos
I think chromosome wasn't written correctly a single time. Closes #40

\author{Juan-J.~Merelo*$^1$, Mario Garc\'ia-Valdez$^2$, Pedro A. Castillo$^1$, Pablo Garc\'ia-S\'anchez$^1$, P. de las Cuevas*$^1$, Nuria Rico$^3$
\thanks{Manuscript submitted for review on \today.}%
\thanks{$^1$Dept. of Computer Architecture and Technology and CITIC University of Granada}%
\thanks{$^2$Dept. of Graduate Studies at Instituto Tecnol\'ogico de Tijuana}%
\thanks{$^3$Dept. of Statistics and Operational Researcha University of Granada}%
\thanks{E-mail addresses: {\tt jmerelo@ugr.es}, {\tt
    mario@tectijuana.edu.mx}, {\tt pacv@ugr.es}, {\tt pablogarcia@ugr.es},{\tt nrico@ugr.es}}%
\thanks{*Corresponding author.}%
}

@@

When developing volunteer based systems, desingers must
also consider the whole {\em social} aspect in the design, with
issues related to security, trust and privacy among others. 
The computing system becomes a {\em techno-social system} \cite{vespignani2009predicting}.  
% (Paloma) Is the social aspect related to the word "safety"? It's not completely clear.
%a {\em social} aspect to the design of a
%computing system that makes it a {\em techno-social system}
In this paper, our intention is to
present the {\sf NodIO} framework and analyze it from two different
angles: the purely technical angle, including the decisions that went
into its features, and the social angle, to account on how the persons
that voluntarily choose to enter the web page running the system
determine its performance and how it can be modeled and, eventually,
predicted.

@@

JavaScript is also compatible with the functional programming paradigm \cite{Cousineau1998,MacLennan1990,Thompson1996}.
According to \cite{swanresearch2015}, the benefit of using this kind of paradigm in
metaheuristics optimization is that it allows an improvement in communicability,
reproducibility, interoperability, automated assembly, knowledge engineering
and efficiency.

@@

In general, a distributed volunteer-based evolutionary computation
system based on the browser is simply a client-server system
whose client is, or it can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. Even so, there might
be some issues with the performance of the language itself; which % (Paloma) I recommend the use of "which" here
 is
why we have made a comparison between JavaScript and other languages
\cite{2015arXiv151101088M}. The results of this work show that JavaScript can be
faster than compiled languages such as Scala and, in any case, it provides
a performance that is comparable to other languages usually employed % (Paloma) Adverb
in evolutionary computation and numerical optimization such as Python. 
% (Paloma) Bit of a too long sentence here. Why not changing the ; for a . ?
Compiled languages such as
Java or C might be faster when % (Paloma) Too many "fact" words. There was another one in the previous sentence.
considering the performance of a single-user, 
but this is more than offset by the computational power of
the spontaneous volunteers we can gather at whim, including people
using their mobile phones or tablets. Together, the performance is several orders of magnitude
higher, which is the objective in this kind of systems.
% (Paloma) This sentence needs re-writing
% "Other compiled languages such as Java or C might be, in fact, faster, but the fact that single-user performance is higher (than what?) is over the offset. This is because the number of spontaneous volunteers we can gather at whim, including people using their mobile phones or tablets, is several orders of magnitude higher, which is the objective in this kind of systems." (?)
 JavaScript is
quite a popular language nowadays, providing compilers from a number
of languages, many of them in their family (like CoffeeScript), but
many other outside, so researchers can, in fact, write their fitness
function in Python, Erlang, Scala, or even Java, and them compile it to
JavaScript. For completness, an experiment comparing the performance of
a JavaScript implementation versus Java and Matlab is included in section
\ref{sec:w2}.

@@

The researcher only has to change a function, which can be written in
JavaScript or in any other language with cross-compilation to
JavaScript \cite{web:compilersjs} to solve a different
problem. In this case the classical Trap function \cite{Ackley1987} has been
used as well as the Rastrigin's floating-point optimization
problem. JavaScript is a functional language and declaring a different 
function and handing at the creation of the algorithm object, called
{\tt Classic}, is the only change needed to work with a different
problem. Let us see how this system addresses the different challenges
outlined in the State of the Art (Section \ref{sec:soa}):
\begin{itemize}
\item {\em Scalability} is provided via the use of a lightweight and
  high-performance, single-threaded, server based in Node.js and
  Express.js. Although this single server is a bottleneck since it
  will eventually saturate, the fact that it runs as a non-blocking single thread
  allows the service of many requests. In fact, a limit in the
  number of simultaneous requests will be reached, but so far it has
  not been found, unlike what we found in our previous systems, DCoR,
  \cite{gecco07:workshop:dcor}, which had a low scalability. 
\item The system is {\em heterogeneous} since it does not need any
  performance, operating system, or even browser requirement as long
  as Javascript is enabled: anyone
  visiting the page, even from mobile devices, can load the algorithm.
  In case the browser does not support HTML5 Web workers (e.g., 
  Windows Phone or some version of Android) or their use
  has been disabled, a basic version of {\sf NodIO} can also be used.
\item {\em Fault tolerance} is always an issue, and in this case, the
  single point of failure would be the server: the system, as such,
  would break down if the server fails. However, the individual
  islands in every browser would continue running, and having access
  to just one of them would allow the local algorithm to proceed. In
  fact, the island does not need the server to run: it runs locally if
  needed, with the only exception that it is obviously unable to
  communicate with the rest of the islands.
\item {\em Adaptiveness} is achieved simply through the autonomous
  operation of every individual island without any synchronization
  mechanism. The islands in the system are, in fact, unaware of each
  other, communicating only through the server. This architecture 
  has been tested in other high demand systems with success, but 
  in the case of {\sf NodIO}, additional experiments are needed to asses 
  the scalability of the communication system. % (Paloma) Is it ok if we talk about scalability when we were talking about adaptiveness? Scalability was also described in the first point.
  %(Mario) Scalability of the communication system in particular?
\item Since the algorithm runs in the browser, {\em safety} is
  achieved through its sandbox mechanisms. The user is thus assured
  that there is no unsafe access either to their local files or even
  to more resources than the browser should be allotted.
\item Running the algorithm is just a matter of loading the page,
  which makes the operation totally {\em anonymous}. For the same
  reason, {\em ease of use} is optimal, being as easy as simply
  clicking on an URL, available to anyone with access to a browser.
\item {\em Reasonable performance} is not ensured. In fact, we should
  make sure that there is a reasonable amount of clients over which
  the performance achieved is better than what you would obtain in
  your own desktop system. If this is not the case, it is a pointless
  academic exercise.
\end{itemize}

@@

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of Internet Protocol addressess (IPs)
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even in this case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen in 1 out of roughly 5 or roughly 20 cases, depending on what baseline
experiment you choose, it will not happen always or even most of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found. This is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. Thus, the user can decide if to continue sharing resources or not.% (Paloma) Feature or disadvantage?
 However, it puts a probably            % (Mario) Feature if you only want to share your CPU a little  
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades
of blue indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequence. From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

@@

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two {\sf NodEO} instances per browser. The sequence diagram shows the
interaction between processes in a time sequence. In this system there are
two kinds of asynchronous messages: HTTP Requests shown as bold arrows between
the client and the server where PUT and GET HTTP Methods are represented by
red and blue arrows respectively. There are also messages between workers and
the main thread locally in the client, which are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em Node.js} This is the server side process which is basically
  the same as used previously, except for some minor log-related
  changes. 
\item {\em Client main thread} The main script creates
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronous HTTP requests to the server are implemented using the
XMLHttpRequest (XHR) object.  
\end{itemize}

%
\begin{figure}[!htb]
\centering
\includegraphics[width=4in]{Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}
\label{fig:system:w2}
\end{figure}

The sequence of events happens as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in Node.js. As part of the
initialization procedure, the sequential number of experiments is initialized and a log
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP
request, then the page is loaded.
\item The main script renders the page and creates the worker instance. In
order to create a dedicated worker instance, another GET request is sent to the
server, then the client receives the second worker implementation script.
Each worker instance executes over time one or more evolutionary algorithms
{\em islands}. Each island starts with a random population and it is assigned a
universally unique identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.
The initial parameters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following the previously
specified parameters. Once the population has been initialized, the EA
generation loop is started. After {\em n} generations the state of the population
is evaluated. The chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not the global best,
the EA loop must continue, but first a chromosome is randomly selected from
the shared population. The current generation number and the best
chromosome are also sent to the client's main thread in order to update the DOM to
display the current state of the island (described next). These messages are
all asynchronous, so the EA loop of
the worker continues even before a message is received by the target. If the best fitness is a global best, the best
chromosome is sent to the server and the loop ends. A message is sent to the
main thread indicating that the best individual has been found.
\item The main thread receives messages from workers as events handled by
a callback function. When an iteration message is received, the page is
updated by displaying a dynamic plot of the generation number and best
fitness, and a representation of the chromosome is also generated.
\item When a global best is received from an island, the current experiment
ends, the experiment number is incremented, and the population array is reset.
\item When the main thread receives a message indicating that a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the start-up cost, only the parameters and population are reset and a new UUID
assigned.
\end{enumerate}
%
\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%

The experiments were run in the same way as before, using several
Twitter announcements throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results is shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, as we considered every worker as a unique IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiments take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that most of the times is better than we would by using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.

%FERGU: a lo mejor algún reviewer se queja de que la comparativa no es justa
%porque comparamos un mono-isla (el baseline) con un multi-isla. Quizás deberíamos
%haber usado un multi-isla en el baseline (aunque sea
%secuencial). Para la próxima.
% se trata de probar que podemos obtener mejores resultados que en una
% implementación local básica. Si usamos una paralela local, ya
% depende del ordenador que tengamos - JJ

@@

\subsection{$F_{15}:\frac{D}{m}$-group Shifted and $m$-rotated Rastrigins Function}
JavaScript has been traditionally implemented as an interpreted
language not designed for the development of high performance
systems. On the other hand, current JavaScript  
virtual machines (VMs) are closing the gap, for instance the Google V8 
is an engine specifically designed for the fast execution of 
large JavaScript applications. In order to increase performance
V8 compiles JavaScript source code directly into machine code when it is first executed. 
There are no intermediate byte codes or an interpreter \cite{Gray:2009:GCM:1610564.1610565}.
JavaScript has gained popularity in server-side development using runtime
environments such as Node.js and is also used in desktop and mobile applications. 
%Mario: Creo que el párrafo de arriba sale sobrando
% (Paloma) Agree, I mean, it adds information but in case you have space problems... is ok if you start the section without it
To further evaluate how a current JavaScript implementation could be used by a
computer scientist to develop complex optimization problems,
a benchmark function, which was provided by the CEC2010 Special Session on
Large-Scale Global Optimization \cite{tang2007benchmark} is used in this section. 
The function is described next.

>>>>

\author{Juan-J.~Merelo*$^1$, Mario Garc\'ia-Valdez$^2$, Pedro A. Castillo$^1$, Pablo Garc\'ia-S\'anchez$^1$, P. de las Cuevas*$^1$, Nuria Rico$^3$
\thanks{Manuscript submitted for review on \today.}%
\thanks{$^1$Dept. of Computer Architecture and Technology and CITIC University of Granada}%
\thanks{$^2$Dept. of Graduate Studies at Instituto Tecnol\'ogico de Tijuana}%
\thanks{$^3$Dept. of Statistics and Operational Research University of Granada}%
\thanks{E-mail addresses: {\tt jmerelo@ugr.es}, {\tt
    mario@tectijuana.edu.mx}, {\tt pacv@ugr.es}, {\tt pablogarcia@ugr.es},{\tt nrico@ugr.es}}%
\thanks{*Corresponding author.}%
}

@@

When developing volunteer based systems, designers must
also consider the whole {\em social} aspect in the design, with
issues related to security, trust and privacy among others. 
The computing system becomes a {\em techno-social system} \cite{vespignani2009predicting}.  
% (Paloma) Is the social aspect related to the word "safety"? It's not completely clear.
%a {\em social} aspect to the design of a
%computing system that makes it a {\em techno-social system}
In this paper, our intention is to
present the {\sf NodIO} framework and analyze it from two different
angles: the purely technical angle, including the decisions that went
into its features, and the social angle, to account on how the persons
that voluntarily choose to enter the web page running the system
determine its performance and how it can be modeled and, eventually,
predicted.

@@

JavaScript is also compatible with the functional programming paradigm \cite{Cousineau1998,MacLennan1990,Thompson1996}.
According to \cite{swanresearch2015}, the benefit of using this kind of paradigm in
metaheuristics optimization is that it allows an improvement in communicability,
reproductibility, interoperability, automated assembly, knowledge engineering
and efficiency.

@@

In general, a distributed volunteer-based evolutionary computation
system based on the browser is simply a client-server system
whose client is, or it can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. Even so, there might
be some issues with the performance of the language itself; which % (Paloma) I recommend the use of "which" here
 is
why we have made a comparison between JavaScript and other languages
\cite{2015arXiv151101088M}. The results of this work show that JavaScript can be
faster than compiled languages such as Scala and, in any case, it provides
a performance that is comparable to other languages usually employed % (Paloma) Adverb
in evolutionary computation and numerical optimization such as Python. 
% (Paloma) Bit of a too long sentence here. Why not changing the ; for a . ?
Compiled languages such as
Java or C might be faster when % (Paloma) Too many "fact" words. There was another one in the previous sentence.
considering the performance of a single-user, 
but this is more than offset by the computational power of
the spontaneous volunteers we can gather at whim, including people
using their mobile phones or tablets. Together, the performance is several orders of magnitude
higher, which is the objective in this kind of systems.
% (Paloma) This sentence needs re-writing
% "Other compiled languages such as Java or C might be, in fact, faster, but the fact that single-user performance is higher (than what?) is over the offset. This is because the number of spontaneous volunteers we can gather at whim, including people using their mobile phones or tablets, is several orders of magnitude higher, which is the objective in this kind of systems." (?)
 JavaScript is
quite a popular language nowadays, providing compilers from a number
of languages, many of them in their family (like CoffeeScript), but
many other outside, so researchers can, in fact, write their fitness
function in Python, Erlang, Scala, or even Java, and them compile it to
JavaScript. For completeness, an experiment comparing the performance of
a JavaScript implementation versus Java and Matlab is included in section
\ref{sec:w2}.

@@

The researcher only has to change a function, which can be written in
JavaScript or in any other language with cross-compilation to
JavaScript \cite{web:compilersjs} to solve a different
problem. In this case the classical Trap function \cite{Ackley1987} has been
used as well as the Rastrigin's floating-point optimization
problem. JavaScript is a functional language and declaring a different 
function and handing at the creation of the algorithm object, called
{\tt Classic}, is the only change needed to work with a different
problem. Let us see how this system addresses the different challenges
outlined in the State of the Art (Section \ref{sec:soa}):
\begin{itemize}
\item {\em Scalability} is provided via the use of a lightweight and
  high-performance, single-threaded, server based in Node.js and
  Express.js. Although this single server is a bottleneck since it
  will eventually saturate, the fact that it runs as a non-blocking single thread
  allows the service of many requests. In fact, a limit in the
  number of simultaneous requests will be reached, but so far it has
  not been found, unlike what we found in our previous systems, DCoR,
  \cite{gecco07:workshop:dcor}, which had a low scalability. 
\item The system is {\em heterogeneous} since it does not need any
  performance, operating system, or even browser requirement as long
  as JavaScript is enabled: anyone
  visiting the page, even from mobile devices, can load the algorithm.
  In case the browser does not support HTML5 Web workers (e.g., 
  Windows Phone or some version of Android) or their use
  has been disabled, a basic version of {\sf NodIO} can also be used.
\item {\em Fault tolerance} is always an issue, and in this case, the
  single point of failure would be the server: the system, as such,
  would break down if the server fails. However, the individual
  islands in every browser would continue running, and having access
  to just one of them would allow the local algorithm to proceed. In
  fact, the island does not need the server to run: it runs locally if
  needed, with the only exception that it is obviously unable to
  communicate with the rest of the islands.
\item {\em Adaptiveness} is achieved simply through the autonomous
  operation of every individual island without any synchronization
  mechanism. The islands in the system are, in fact, unaware of each
  other, communicating only through the server. This architecture 
  has been tested in other high demand systems with success, but 
  in the case of {\sf NodIO}, additional experiments are needed to asses 
  the scalability of the communication system. % (Paloma) Is it ok if we talk about scalability when we were talking about adaptiveness? Scalability was also described in the first point.
  %(Mario) Scalability of the communication system in particular?
\item Since the algorithm runs in the browser, {\em safety} is
  achieved through its sandbox mechanisms. The user is thus assured
  that there is no unsafe access either to their local files or even
  to more resources than the browser should be allotted.
\item Running the algorithm is just a matter of loading the page,
  which makes the operation totally {\em anonymous}. For the same
  reason, {\em ease of use} is optimal, being as easy as simply
  clicking on an URL, available to anyone with access to a browser.
\item {\em Reasonable performance} is not ensured. In fact, we should
  make sure that there is a reasonable amount of clients over which
  the performance achieved is better than what you would obtain in
  your own desktop system. If this is not the case, it is a pointless
  academic exercise.
\end{itemize}

@@

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of Internet Protocol addresses (IPs)
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even in this case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen in 1 out of roughly 5 or roughly 20 cases, depending on what baseline
experiment you choose, it will not happen always or even most of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found. This is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. Thus, the user can decide if to continue sharing resources or not.% (Paloma) Feature or disadvantage?
 However, it puts a probably            % (Mario) Feature if you only want to share your CPU a little  
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades
of blue indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequency. From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

@@

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two {\sf NodEO} instances per browser. The sequence diagram shows the
interaction between processes in a time sequence. In this system there are
two kinds of asynchronous messages: HTTP Requests shown as bold arrows between
the client and the server where PUT and GET HTTP Methods are represented by
red and blue arrows respectively. There are also messages between workers and
the main thread locally in the client, which are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em Node.js} This is the server side process which is basically
  the same as used previously, except for some minor log-related
  changes. 
\item {\em Client main thread} The main script creates
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronous HTTP requests to the server are implemented using the
XMLHttpRequest (XHR) object.  
\end{itemize}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=4in]{Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}
\label{fig:system:w2}
\end{figure}
%
\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline
\end{tabular}
\end{center}
\end{table*}

The sequence of events happens as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in Node.js. As part of the
initialization procedure, the sequential number of experiments is initialized and a log
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP
request, then the page is loaded.
\item The main script renders the page and creates the worker instance. In
order to create a dedicated worker instance, another GET request is sent to the
server, then the client receives the second worker implementation script.
Each worker instance executes over time one or more evolutionary algorithms
{\em islands}. Each island starts with a random population and it is assigned a
universally unique identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.
The initial parameters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following the previously
specified parameters. Once the population has been initialized, the EA
generation loop is started. After {\em n} generations the state of the population
is evaluated. The chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not the global best,
the EA loop must continue, but first a chromosome is randomly selected from
the shared population. The current generation number and the best
chromosome are also sent to the client's main thread in order to update the DOM to
display the current state of the island (described next). These messages are
all asynchronous, so the EA loop of
the worker continues even before a message is received by the target. If the best fitness is a global best, the best
chromosome is sent to the server and the loop ends. A message is sent to the
main thread indicating that the best individual has been found.
\item The main thread receives messages from workers as events handled by
a callback function. When an iteration message is received, the page is
updated by displaying a dynamic plot of the generation number and best
fitness, and a representation of the chromosome is also generated.
\item When a global best is received from an island, the current experiment
ends, the experiment number is incremented, and the population array is reset.
\item When the main thread receives a message indicating that a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the start-up cost, only the parameters and population are reset and a new UUID
assigned.
\end{enumerate}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The red dot and line show the average and standard
  deviation, the shade of blue the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
\end{figure}

The experiments were run in the same way as before, using several
Twitter announcements throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results is shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, as we considered every worker as a unique IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiments take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that most of the times is better than we would by using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value (GEV) and Weibull distribution of
  the number of {\tt PUT}s per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.83719  &  25.962  $\pm$ 1.22015 & 1.242   $\pm$
0.04746 \\
Weibull & ND & 75.11457519 $\pm$ 2.97470251  & 0.69714410 $\pm$ 0.01383976 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%

@@

\subsection{$F_{15}:\frac{D}{m}$-group Shifted and $m$-rotated Rastrigin's Function}
JavaScript has been traditionally implemented as an interpreted
language not designed for the development of high performance
systems. On the other hand, current JavaScript  
virtual machines (VMs) are closing the gap, for instance the Google V8 
is an engine specifically designed for the fast execution of 
large JavaScript applications. In order to increase performance
V8 compiles JavaScript source code directly into machine code when it is first executed. 
There are no intermediate byte codes or an interpreter \cite{Gray:2009:GCM:1610564.1610565}.
JavaScript has gained popularity in server-side development using runtime
environments such as Node.js and is also used in desktop and mobile applications. 
%Mario: Creo que el párrafo de arriba sale sobrando
% (Paloma) Agree, I mean, it adds information but in case you have space problems... is ok if you start the section without it
To further evaluate how a current JavaScript implementation could be used by a
computer scientist to develop complex optimization problems,
a benchmark function, which was provided by the CEC2010 Special Session on
Large-Scale Global Optimization \cite{tang2007benchmark} is used in this section. 
The function is described next.
